---
title: "Bioconductor Exercises Part 3"
author: "MRC LMS Bioinformatics Core"
date: "19 September 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

----
```{r echo=FALSE, eval=TRUE, warning=FALSE}

```


# Solutions

#### 1. Check what kind of annotation data can be extracted using `org.Hs.eg.db` package
```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
library("org.Hs.eg.db")
columns(org.Hs.eg.db)
```

#### 2. Print few gene names from `org.Hs.eg.db`
```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
head(keys(org.Hs.eg.db,keytype="GENENAME"))
```

#### 3. Retrieve gene name and Ensembl gene identifiers for “HEBP2” and “PRND” from `org.Hs.eg.db`

```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE, warning=FALSE}
select(org.Hs.eg.db,keys=c("HEBP2","PRND"),keytype="SYMBOL",
       columns=c("GENENAME","ENSEMBL"))
```

#### 4. How many annotation datasets available in Ensembl Biomart?
```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE, warning=FALSE}
library("biomaRt")
ensembl <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2015.archive.ensembl.org") # select ensembl
ens_datasets <- listDatasets(ensembl) # list datasets 
head(ens_datasets)
dim(ens_datasets)
```

```{r, echo=F, eval=F, message=FALSE, collapse=TRUE, warning=FALSE}
ensembl.19 <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2015.archive.ensembl.org")
```

#### 5. Retrieve genomic coordinates for human genes from Ensembl biomart and build a GRanges object.
  + Subset the above GRanges object to include only protein coding genes
  + Subset the GRanges object again with genes in main chromsomes (1-22,X,Y)
  + Create another GRanges object with genes in chr1:1544000-2371000
      
      + Tips:
      + You can select main chromosomes and "protein coding" genes by using appropriate filter and value.
      + Search for "biotype" in available filters using `grep()`
      + Run `filterOptions("biotype",selectedmart)` to see the accepted values for "biotype" filter
      + When multiple filters specified, "values" argument should be a list of vectors; each vector corresponds to each specified filter.
      + Annotation fields to retrieve: "chromosome_name", "start_position", "end_position","ensembl_gene_id", "strand", "external_gene_name"
      + Before creating GRanges object, add "chr" prefix to chromosome using `paste` function, Ex: change 1 to chr1 (required for next task)

```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE, warning=FALSE}
ens_human <- useDataset("hsapiens_gene_ensembl",mart=ensembl) # select human dataset
ens_human_Attr <- listAttributes(ens_human) # list available annotation
ens_human_filters <- listFilters(ens_human) # list available filters
availFilters <- filterOptions("biotype",ens_human) # Displays accepted values for "biotype"

hg19Gene <- getBM(
          attributes = c("chromosome_name","start_position","end_position",
                         "ensembl_gene_id","strand","external_gene_name"), 
          filter=c("chromosome_name","biotype"),
          values=list(c(1:22,"X","Y"),"protein_coding"), mart=ens_human)
head(hg19Gene)
```

Now create GRanges object using the above data frame. 
```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
library(GenomicRanges)

# add 'chr' prefix to chromosome name
hg19Gene$chromosome_name <- paste("chr",hg19Gene$chromosome_name,sep="")

hg19Gene.GR <- GRanges(seqnames=hg19Gene$chromosome_name,
                       ranges=IRanges(start=hg19Gene$start_position,end=hg19Gene$end_position),
                       strand=ifelse(hg19Gene$strand==1,"+","-"),
                       EnsemblID=hg19Gene$ensembl_gene_id,
                       Symbol=hg19Gene$external_gene_name)
hg19Gene.GR
```

Filter the above GRanges object for genes in chr1:1544000-2371000
```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
chr1genes <- hg19Gene.GR[seqnames(hg19Gene.GR)=="chr1" & 
                         start(hg19Gene.GR) > 1544000 & 
                           end(hg19Gene.GR) < 2371000]
head(chr1genes)

# alternate solution
chr1genes <- subset(hg19Gene.GR,start>1544000 & end<2371000 & seqnames=="chr1")
```

#### 6. Retrieve 200 bp upstream promoter sequences for the given gene symbols `AQP1, ASNSP2, KPNA2, FRMD4A, NSUN5, VAC14` from Ensembl human biomart.
    + Tips: 
    + Read documentation for `getSequence`
    + Use `type="hgnc_symbol"` and `seqType="coding_gene_flank"`

```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE, warning=FALSE}
symbols <- c("AQP1", "ASNSP2", "KPNA2", "FRMD4A", "NSUN5", "VAC14")
ensembl <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2015.archive.ensembl.org", 
                   dataset="hsapiens_gene_ensembl") # select ensembl

seq <- getSequence(id=symbols, type="hgnc_symbol",
                  seqType="coding_gene_flank", upstream=200, mart = ensembl)
```


#### 7. Retrieve the transcript coordinates for genes as GRangesList from `TxDb.Hsapiens.UCSC.hg19.knownGene` (install it from Bioconductor if required)
```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
# source("http://bioconductor.org/biocLite.R")
# biocLite("TxDb.Hsapiens.UCSC.hg19.knownGene")
library("TxDb.Hsapiens.UCSC.hg19.knownGene")
hg19txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
TranscrtipsByGene <- transcriptsBy(hg19txdb,by="gene") # inspect the output
```

<br>

Alternate solution

```{r, echo=TRUE, eval=F, message=FALSE, collapse=TRUE}
library(GenomicFeatures)
hg19txdb <- makeTxDbFromUCSC(genome = "hg19", tablename = "knownGene")
TranscrtipsByGene <- transcriptsBy(hg19txdb,by="gene") # inspect the output
```

#### 8. Retrive exon coordiantes for genes from `TxDb.Hsapiens.UCSC.hg19.knownGene`
```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
ExonsByGene <- exonsBy(hg19txdb,by="gene") # inspect the output
```

