---
title: "Bioconductor Exercises Part 1"
author: "MRC LMS Bioinformatics Core"
date: "19 September 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

----
```{r echo=FALSE, eval=TRUE, warning=FALSE}
setwd("/Volumes/bioinfomatics$/yfwang/CBW/Bioconductor_2019/course")
```


# Solutions

#### 1. Import human hg19 gene coordiantes from "hg19Genes.txt and create a GRanges of transcription start sites (1 bp range).

    + Tip: How to identify TSS for genes in forward/reverse strand?

```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
library(GenomicRanges)
hg19Gene <- read.table("./data/hg19Genes.txt",sep="\t",header=T)

hg19Gene$TSS <- ifelse(hg19Gene$strand=="+",hg19Gene$start,hg19Gene$end)

hg19TSS <- GRanges(seqnames=hg19Gene$chr,
                       ranges=IRanges(start=hg19Gene$TSS,end=hg19Gene$TSS),
                       strand=hg19Gene$strand,
                       EnsemblID=hg19Gene$ensID,
                       Symbol=hg19Gene$GeneSym)
hg19TSS
```


#### 2. Create a GRanges object of human promoters with TSS ± 2000bp.

    + Tip: Read the documentation for `promoters` function.
    
```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
# First create GRanges object with gene coordinates
hg19Gene.GR <- GRanges(seqnames=hg19Gene$chr,
                       ranges=IRanges(start=hg19Gene$start,end=hg19Gene$end),
                       strand=hg19Gene$strand,
                       EnsemblID=hg19Gene$ensID,
                       Symbol=hg19Gene$GeneSym)

# Identifying promoters
hg19Promoters <- promoters(hg19Gene.GR,upstream=2000,downstream=2000)
hg19Promoters
```

#### 3. Import ELF1 binding sites in K562 cell from Encode (ELF1_K562.bed) and create GRanges object. 

    + Tips: Import the ELF1 binding sites using `import.bed()` function from `rtracklayer` package
    + Find ELF1 binding sites overlap with promoters (TSS ± 1kb) using `findOverlaps` and `subsetByOverlaps` (Remember BED format uses 0-based coordinates)

```{r, echo=TRUE, eval=TRUE, message=FALSE, collapse=TRUE}
library("rtracklayer")
ELF1 <- read.table("./data/ELF1_K562.bed",sep="\t",header=F)
ELF1GR <- GRanges(seqnames=ELF1$V1, IRanges(start=ELF1$V2+1,end=ELF1$V3))
ELF1GR_A <- import.bed("./data/ELF1_K562.bed")


# ELF1 binding sites overlap with promoters using `findOverlaps`
hg19Promoters <- promoters(hg19Gene.GR,upstream=1000,downstream=1000)
ELF1GR <- reduce(ELF1GR) # merging overlapping peaks
ELF1overlap <- findOverlaps(ELF1GR,hg19Promoters,ignore.strand=T)
ELF1overlap.m <- as.matrix(ELF1overlap)
ELF1_promoters <- ELF1GR[ELF1overlap.m[,"queryHits"],]
ELF1_promoters

# ELF1 binding sites overlap with promoters using `subsetByOverlaps`
ELF1_promoters1 <- subsetByOverlaps(ELF1GR,hg19Promoters)
ELF1_promoters1
```

<b>Note the differences in the outputs!</b>