---
title: "Bioconductor Exercises Part 2"
author: "MRC LMS Bioinformatics Core"
date: "26 September 2018"
output: html_document
---

----
```{r echo=FALSE, eval=TRUE, warning=FALSE}

```


# Solutions

* Download the following BAM and index files (*.bai) (ENCODE data - ChIP-Seq of CTCF in Ag04449 human fibroblast cells)
    - http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeUwTfbs/wgEncodeUwTfbsAg04449CtcfStdAlnRep1.bam
    - http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeUwTfbs/wgEncodeUwTfbsAg04449CtcfStdAlnRep1.bam.bai
* Use `readGAlignments` to read the bam. Construct an `ScanBamParam` object that accepts only aligned reads, passing quality control and not duplicates.
* Compute genome wide coverage using `coverage` function
* Export coverage as bigWig using `export.bw()` from rtracklayer package and visualise it using IGV
* Compute number of reads overlapping with hg19 promoters (TSS ± 1kb) and export the results as text file.

```{r, echo=TRUE, eval=FALSE, message=FALSE, collapse=TRUE}
library("GenomicAlignments")
library("rtracklayer")
bamFile <- "wgEncodeUwTfbsAg04449CtcfStdAlnRep1.bam"
flag <- scanBamFlag()
param <- ScanBamParam(
    flag=scanBamFlag(isUnmappedQuery=FALSE, isDuplicate=FALSE, isNotPassingQualityControls=FALSE)
    )

# Read the BAM
CTCF <- readGAlignments(bamFile,param=param)

# Generate the genomic coverage and inspect the output
CTCFCov <- coverage(CTCF)

# Export as bigWig
export.bw(CTCFCov, "Ag04449_CTCF.bw")

# Identifying hg19 promoters
hg19Gene <- read.table("hg19Genes.txt",sep="\t",header=T)
hg19Gene.GR <- GRanges(seqnames=hg19Gene$chr,
                       ranges=IRanges(start=hg19Gene$start,end=hg19Gene$end),
                       strand=ifelse(hg19Gene$strand==1,"+","-"),
                       EnsemblID=hg19Gene$ensID,
                       Symbol=hg19Gene$GeneSym)
hg19Promoters <- promoters(hg19Gene.GR,upstream=2000,downstream=2000)

# Reads overlapping with promoters
CTCFCounts <- countOverlaps(hg19Promoters,CTCF)

# Add CTCF counts as elementMetadata to hg19Promoters object
mcols(hg19Promoters)$CTCF <- CTCFCounts

# Export the results as text file
hg19Promoters.df <- as.data.frame(hg19Promoters)
write.table(hg19Promoters.df,"hg19Promoters_CTCF.txt",sep="\t",row.names=F)

```

